---
output: pdf_document
---

# Conclusion {.unnumbered}
  \setcounter{chapter}{4}
	\setcounter{section}{0}

## Concluding Thoughts

While Independent Component Analysis can be highly effective on a wide range of datasets, its ambiguities regarding particular results and relatively difficulty of interpretation explain the lack of widespread usage by the Statistical community at large.  Even if the key model criteria (independence and non-Gaussianity) are met, choosing the correct number of components and making sense of the output requires a deep understanding of the data.  In their study on parallel time series financial data, Kiviluoto and Oja note how the method successfully revealed underlying factors in the retail chain analyzed but also how ICA modeling calls for "inspection from a domain expert." [pg. 3,@kimmo1998] 

Perhaps the best example of this need for experts to analyze results would be the work done with the ICA model on brain imaging data like EEG and MEG.  In many ways, ICA has proven to be a valuable tool for neuroscience researchers, but this application would not be possible with analysis by statisticians alone. Although successful understanding of results in this case may always be limited to experts, new ICA algorithms with more explicit outputs could potentially be developed to improve breadth of its application.  



 


<!--
If you feel it necessary to include an appendix, it goes here.
-->

\appendix

\singlespacing

#

This first appendix includes all of the R chunks of code that were hidden throughout the document (using the `include = FALSE` chunk tag) to help with readability and/or setup.

#### In the main Rmd file:

```{r ref.label = 'include_packages', results = 'hide', echo = TRUE}
```

#### In [](#ref_labels):

```{r ref.label = 'include_packages_2', results = 'hide', echo = TRUE}
```

# The Second Appendix, for Fun

